\OurTool{}: Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles

### 2020-07-02 IPANEMAP-2019.tex
Updates from Overleaf

Reagents used within \RevA{selective 2' -hydroxyl acylation analyzed by primer extension (SHAPE)~\cite{Wilkinson2006} protocols} represent a popular class of probes. They react with the 2’-hydroxyl of flexible ribose~\cite{McGinnis2012}, although the exact properties observed by SHAPE remain the object of ongoing investigations~\cite{McGinnis2012,Sexton2017,Hurst2018,Mlynsky2018,Frezza2019,Busan2019}.
As ribose flexibility is proportional to the degree of freedom of the nucleotide, it is assumed that SHAPE reagents discriminate nucleotides involved in stable interactions. Different SHAPE reagents are endowed with different dynamics, that can help differentiate Watson-Crick base pairs from more dynamic tertiary contacts~\cite{Gherghe2008,Steen2012,Rice2014,Busan2019}. Other chemical probes, harbouring different chemical reactivities, have been developed. Amongst the most popular, DiMethyl Sulfate (DMS) is a small molecule that methylates Adenines and Cytosines if not involved in a hydrogen bond, while   CMCT reacts with the Watson-Crick face of unpaired Guanosines and Uracils~\cite{Ehresmann1987,Brunel2000}. These two reagents not only reveal Watson-Crick base pairing, but also other types of contacts involving the same edge. The diversity of probes\RevA{, some of which are usable \emph{in vivo},} not only increases coverage of the different positions and structural contexts, \RevA{but also provides different qualitative information. \RevA{Modeling can also benefit from a joint analysis of reactivity profiles of single-point mutants, assuming structural homology~\cite{Cordero2012a}}. Integration of multiple sources of probing to improve structure prediction  has thus been widely used since the very early days of RNA structure probing~\cite{Moazed1986, Ehresmann1987, Romaniuk1988, Butcher1994, Brunel2000,  Cordero2015, Somarowthu2015, Gross2017} but, to the best of our knowledge, has never been fully automated.}

@@

The targeted number of clusters is a critical parameter of the \CL{} algorithm. It should, at the same time, remain small enough to ensure reproducibility, while being sufficiently large to discriminate outliers and ensure consistency within each cluster. We determine an \Def{optimal number of clusters} $\NumClust$ using an iterative heuristics, gradually increasing the number of clusters until a significantly-populated cluster is split into two similar clusters, or a poorly populated (outlier) cluster is created.
Namely, our iterative heuristic consists in running \CL over an increasing number $\NumClust$ of clusters, starting with $\NumClust=2$, until the following \Def{stopping criterion} is met: 1) Two significantly populated clusters have associated centroids which are highly similar; or 2) Centroid structures of significantly populated clusters from the previous iteration are highly similar to those of the current iteration.}

\paragraph{Filtering the promising conformer(s).} Finally, we \Def{elect the most promising cluster(s)}, and return their centroid(s). While the final number of clusters may potentially be large, only a handful of clusters are expected to represent structures that \RevA{are} both stable, and supported by a large number of conditions. The remaining clusters are indeed probably artifacts of the clustering method, but nevertheless useful to filter out \emph{noisy} structures.

@@

\paragraph{Cordero dataset.} 
In order to test the integration of different probing sources, we consider a dataset introduced in Cordero\etal\cite{Cordero2012}, consisting of $6$ RNAs with known structures, for which reactivities are available for three reagents: {\tt NMIA-\SH, DMS} and {\tt CMCT}. 
Probing data were downloaded from the RMDB~\citep{Cordero2012a} on July 2017. In the RMDB, reactivity scores are reported for all nucleotides, including those that are not expected to react with a given reagent. Thus, for the DMS (resp. CMCT) probing, we restricted reactivities to positions featuring nucleotides {\sf A} and {\sf C} (resp. {\sf G} and {\sf U}), setting the reactivity to 0 for other positions. This allowed to decrease the noise generated by reactivities associated with non-targeted nucleotides, leading to more accurate predictions (data not shown).

@@

As a preliminary experiment, we verified that \didy{} generally adopts a single global architecture. To this end, we subjected \didy{} to a standard denaturation protocol (80\degree{}C for 2 min in H$_\text{2}$O, addition of 40 mM of HEPES at 7.5 pH, 100 mM of KCl, 5 mM of MgCl$_\text{2}$, followed by 10 min at room temperature, and 10 min at 37\degree{}C), and observed the production of a single band on a non-denaturing PAGE, strongly suggesting the adoption of a single conformation.

@@

\RevA{
\subsection*{Assessment of statistical significance}
Following Xu\etal\cite{Xu2012}, we assessed the statistical significance of observed changes in predictive performances using a two-tailed Student $t$-test (unequal variance) with usual significance threshold $\alpha=5\%$. When comparing performances within a given dataset, we used the paired version of the test.

@@

\subsection*{\OurTool{} \RevB{performs comparably to} state-of-the-art predictive methods when informed by a single profile}
%Mono-probing Mathews (Hadjim)
%* On est meilleur sur la version courante de RSample
%* Multi conformation
We assessed the predictive capacities of \OurTool in a classic setting where a single probing condition is available, and compared its performances against \Software{RSample} software recently introduced by Spacic\etal\cite{Spasic2017}, \Software{RSample} relies on a sampling/clustering method developed independently from the current work. It was shown to perform \RevA{comparably as} a comprehensive collection of state-of-the-art methods in probing-guided structure prediction, including \Software{RME}~\citep{Wu2015}, \Software{RNAprob}~\citep{Deng2016}, \Software{RNAprobing}~\citep{Washietl2012}, \Software{RNAsc}~\citep{Zarringhalam2012} and the \Software{fold} utility from the \Software{RNAStructure} suite~\citep{Reuter2010}.
We considered the Hajdin dataset~\cite{Hajdin2013}, which consists of 24 RNAs of lengths ranging from 47 to 500 nts, all believed to fold into a unique documented conformation.

@@

\RevB{%
\OurTool{} appears to perform comparably, or even slightly favorably, to \Software{RSample}, reporting more accurate structures than its competitor for roughly two thirds of the tested RNAs, while averaging $80.1\%$ GM. However, the  improvement is only of around 1.5\% on average over the \citet{Hajdin2013} dataset, an improvement that is no longer significant according to a two-tailed paired T-test.
}

@@

\paragraph{Considering three conditions yields improvement over the average performance of the triplet.}
Increasing the number of conditions to three boosted the average MCC to 72.3\% and even reached 76.5\% over triplets that do not contain conditions from cluster \Bull{B}. The MCC gain over the worst condition in the triplet reached 9.1\% on average.  Predictions were also more consistently good, with a median MCC of \%73.4. Disregarding conditions in cluster \Bull{B} did stabilize the quality of prediction (stdev 8.4\% $\to$ 5.5\%), while increasing the average MCC to 77.4\%.

@@

To the best of our knowledge, \OurTool currently represents the only available method for a joint automated consideration of multiple probing profiles. Surprisingly, it also appears to produce the most accurate secondary structure models when informed by a single condition, as demonstrated by our reanalysis on the Hajdin\etal\cite{Hajdin2013} benchmark. We attribute this unexpected result to the adoption of a sampling/clustering approach, \RevA{which appears to eliminate spurious outliers from the conformational space even when a single conformation dominates.}
\RevA{In this context, the good performance of \OurTool in comparison with \Software{RSample}~\cite{Spasic2017} may appear surprising. Indeed, while independently developed and used~\cite{Deforges2017}, both methods share a common sampling/clustering scheme. However, \Software{RSample}~\cite{Spasic2017} reverts to a MEA~\cite{Lu2009} computation whenever a single conformation is suspected. The MEA structure coincides with the centroid of the whole Boltzmann ensemble, suggesting that the increased performances of \OurTool could be at least partially attributed to the use of sampling, and to our adaptive choice of the number of clusters.}
Moreover, even in cases where a single dominant conformation is expected, considering multiple clusters seems to create more opportunities for discarding noisy random unstable conformations. Such conformations naturally occur in a Boltzmann-distributed subset of structures, and may end up \emph{polluting} the centroid structures by artificially inflating the cluster diversity. 
%The impact on \Software{RSample} predictions of such a \emph{noise} seems worsened by the fact that it returns the MEA, the centroid of the whole ensemble, whenever a dominant conformation is detected.

Our analyses reveal that integrating multiple profiles of at least three conditions allows to mitigate the effect of outlier conditions and, at times, to produce better secondary structure models than the one inferred only from the most accurate profile. A comparison of the joint performances to the average appears fair since no single reagent appears to systematically dominate in term of predictive capacity. It follows that a modeler cannot objectively favor \emph{a priori} a probing condition/reagent~\cite{Yu2018}. In other words, multiple experimental probing profiles, even combined in a pairwise fashion, \RevA{can be used jointly to mitigate the empirical risk of a misprediction}.

>>>>

Reagents used within \RevA{selective 2' -hydroxyl acylation analyzed by primer extension (SHAPE)~\cite{Wilkinson2006} protocols} represent a popular class of probes. They react with the 2’-hydroxyl of flexible ribose~\cite{McGinnis2012}, although the exact properties observed by SHAPE remain the object of ongoing investigations~\cite{McGinnis2012,Sexton2017,Hurst2018,Mlynsky2018,Frezza2019,Busan2019}.
As ribose flexibility is proportional to the degree of freedom of the nucleotide, it is assumed that SHAPE reagents discriminate nucleotides involved in stable interactions. Different SHAPE reagents are endowed with different dynamics that can help differentiate Watson-Crick base pairs from more dynamic tertiary contacts~\cite{Gherghe2008,Steen2012,Rice2014,Busan2019}. Other chemical probes, harbouring different chemical reactivities, have been developed. Amongst the most popular, DiMethyl Sulfate (DMS) is a small molecule that methylates Adenines and Cytosines if not involved in a hydrogen bond, while   CMCT reacts with the Watson-Crick face of unpaired Guanosines and Uracils~\cite{Ehresmann1987,Brunel2000}. These two reagents not only reveal Watson-Crick base pairing, but also other types of contacts involving the same edge. The diversity of probes\RevA{, some of which are usable \emph{in vivo}\RevB{~\cite{Zaug1995,Paillart2004}},} not only increases coverage of the different positions and structural contexts, \RevA{but also provides different qualitative information. \RevA{Modeling can also benefit from a joint analysis of reactivity profiles of single-point mutants, assuming structural homology~\cite{Cordero2012a}}. Integration of multiple sources of probing to improve structure prediction has thus been widely used since the very early days of RNA structure probing~\cite{Moazed1986, Ehresmann1987, Romaniuk1988, Butcher1994, Brunel2000,  Cordero2015, Somarowthu2015, Gross2017} but, to the best of our knowledge, has never been fully automated \RevB{within a soft constraint framework}~\cite{Lorenz2016}.}

@@

The targeted number of clusters is a critical parameter of the \CL{} algorithm. It should, at the same time, remain small enough to ensure reproducibility, while being sufficiently large to discriminate outliers and ensure consistency within each cluster. We determine an \Def{optimal number of clusters} $\NumClust$ using an iterative heuristic, gradually increasing the number of clusters until a significantly-populated cluster is split into two similar clusters, or a poorly populated (outlier) cluster is created.
Namely, our iterative heuristic consists in running \CL over an increasing number $\NumClust$ of clusters, starting with $\NumClust=2$, until the following \Def{stopping criterion} is met: 1) Two significantly populated clusters have associated centroids which are highly similar; or 2) Centroid structures of significantly populated clusters from the previous iteration are highly similar to those of the current iteration.}

\paragraph{Filtering the promising conformer(s).} Finally, we \Def{select the most promising cluster(s)}, and return their centroid(s). While the final number of clusters may potentially be large, only a handful of clusters are expected to represent structures that \RevA{are} both stable, and supported by a large number of conditions. The remaining clusters are indeed probably artifacts of the clustering method, but nevertheless useful to filter out \emph{noisy} structures.

@@

\paragraph{Cordero dataset.} 
Probing data were downloaded from the RMDB~\citep{Cordero2012a} on July 2017. In the RMDB, reactivity scores are reported for all nucleotides, including those that are not expected to react with a given reagent. Thus, for the DMS (resp. CMCT) probing, we restricted reactivities to positions featuring nucleotides {\sf A} and {\sf C} (resp. {\sf G} and {\sf U}), setting the \RevB{pseudo-energy term to 0 kcal.mol$^{\text{-1}}$} for other positions. This allowed to decrease the noise generated by reactivities associated with non-targeted nucleotides, leading to more accurate predictions (data not shown).

@@

As a preliminary experiment, we verified that \didy{} generally adopts a single global architecture. To this end, we subjected \didy{} to a \RevB{standard denaturation/renaturation protocol} (80\degree{}C for 2 min in H$_\text{2}$O, addition of 40 mM of HEPES at 7.5 pH, 100 mM of KCl, 5 mM of MgCl$_\text{2}$, followed by 10 min at room temperature, and 10 min at 37\degree{}C), and observed the production of a single band on a non-denaturing PAGE, strongly suggesting the adoption of a single conformation.

@@

\RevA{
\subsection*{Assessment of statistical significance}
Following Xu\etal\cite{Xu2012}, we assessed the statistical significance of observed changes in predictive performances using a two-tailed Student $t$-test (unequal variance) with \RevB{usual type I error rate} $\alpha=5\%$. When comparing performances within a given dataset, we used the paired version of the test.

@@

\subsection*{\OurTool{} \RevB{performs comparably to} state-of-the-art predictive methods when informed by a single profile}
%Mono-probing Mathews (Hadjim)
%* On est meilleur sur la version courante de RSample
%* Multi conformation
We assessed the predictive capacities of \OurTool in a classic setting where a single probing condition is available, and compared its performances against \Software{RSample} software recently introduced by Spasic\etal\cite{Spasic2017}, \Software{RSample} relies on a sampling/clustering method developed independently from the current work. It was shown to perform \RevA{comparably as} a comprehensive collection of state-of-the-art methods in probing-guided structure prediction, including \Software{RME}~\citep{Wu2015}, \Software{RNAprob}~\citep{Deng2016}, \Software{RNAprobing}~\citep{Washietl2012}, \Software{RNAsc}~\citep{Zarringhalam2012} and the \Software{fold} utility from the \Software{RNAStructure} suite~\citep{Reuter2010}.
We considered the Hajdin dataset~\cite{Hajdin2013}, which consists of 24 RNAs of lengths ranging from 47 to 500 nts, all believed to fold into a unique documented conformation.

@@

\RevB{%
\OurTool{} appears to perform similarly, or even slightly favorably in comparison with \Software{RSample}. It averages 80.1\% GM, and reports more accurate structural models than its competitor for 14 RNAs in the Hajdin dataset \cite{Hajdin2013}, returns structures equally as good on 2 RNAs, and is dominated by \Software{RSample} for 8 RNAs. However, its modest average improvement of approximately 1.5\% over the dataset is not statistically significant, according to a two-tailed paired T-test. Those results support the notion that \OurTool represents a competitive alternative to state-of-the-art methods for single reactivity profiles.
}

@@

\paragraph{Considering three conditions yields improvement over the average of mono conditions within the same triplet.}
Increasing the number of conditions to three boosted the average MCC to 72.3\% and even reached 76.5\% over triplets that do not contain conditions from cluster \Bull{B}. The MCC gain over the worst condition in the triplet reached 9.1\% on average.  Predictions were also more consistently good, with a median MCC of \%73.4. Disregarding conditions in cluster \Bull{B} did stabilize the quality of prediction (stdev 8.4\% $\to$ 5.5\%), while increasing the average MCC to 77.4\%.

@@

To the best of our knowledge, \OurTool currently represents the only available method for a joint automated consideration of multiple probing profiles. 
%Surprisingly, it also appears to produce the most accurate secondary structure models when informed by a single condition, as demonstrated by our reanalysis on the Hajdin\etal\cite{Hajdin2013} benchmark. We attribute this unexpected result to the adoption of a sampling/clustering approach, \RevA{which appears to eliminate spurious outliers from the conformational space even when a single conformation dominates.}
%\RevA{In this context, the good performance of \OurTool in comparison with \Software{RSample}~\cite{Spasic2017} may appear surprising. Indeed, while independently developed and used~\cite{Deforges2017}, both methods share a common sampling/clustering scheme. However, \Software{RSample}~\cite{Spasic2017} reverts to a MEA~\cite{Lu2009} computation whenever a single conformation is suspected. The MEA structure coincides with the centroid of the whole Boltzmann ensemble, suggesting that the increased performances of \OurTool could be at least partially attributed to the use of sampling, and to our adaptive choice of the number of clusters.}
%Moreover, even in cases where a single dominant conformation is expected, considering multiple clusters seems to create more opportunities for discarding noisy random unstable conformations. Such conformations naturally occur in a Boltzmann-distributed subset of structures, and may end up \emph{polluting} the centroid structures by artificially inflating the cluster diversity. 
%The impact on \Software{RSample} predictions of such a \emph{noise} seems worsened by the fact that it returns the MEA, the centroid of the whole ensemble, whenever a dominant conformation is detected.
Our analyses reveal that integrating multiple profiles of at least three conditions allows to mitigate the effect of outlier conditions and, at times, to produce better secondary structure models than the one inferred only from the most accurate profile. A comparison of the joint performances to the average appears fair since no single reagent appears to systematically dominate in term of predictive capacity. It follows that a modeler cannot objectively favor \emph{a priori} a probing condition/reagent~\cite{Yu2018}. In other words, multiple experimental probing profiles, even combined in a pairwise fashion, \RevA{can be used jointly to mitigate the empirical risk of a misprediction}.
